---
title: "SDPD Annual Service Call Count by Beat"
author: "Sean Jackson"
date: "8/10/2022"
output: 
  html_document:
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
## initial settings
knitr::opts_chunk$set(
  comment = NA,
  echo = FALSE,  # Hide code output
  error = TRUE,
  cache = FALSE,
  message = FALSE,
  dpi = 96,
  warning = FALSE
)

## width to use when printing tables etc.
options(
  width = 250,
  scipen = 100,
  max.print = 5000,
  stringsAsFactors = FALSE
)
```

<style>
.table {
  width: auto;
}
ul, ol {
  padding-left: 18px;
}
pre, code, pre code {
  overflow: auto;
  white-space: pre;
  word-wrap: normal;
  background-color: #ffffff;
}
</style>


```{r packages, include=FALSE}
library(tidyverse)
library(lubridate)
library(leaflet)
library(geojsonio)
library(sf)
library(viridis)
library(ggExtra)
library(shiny)
```


## San Diego Police Department Total Calls for Service by Beat

Leaflet map displays each SDPD beat and the total number of calls for service from the public for each year from 2016 to 2024. Data are directly loaded from the San Diego Open Data portal into the RMD doc where it is cleaned and the map generated. 

## R Code

```{r data loading}
# Load datasets dynamically and combine
years <- 2022:2024
pd_data_list <- lapply(years, function(year) {
  url <- paste0("https://seshat.datasd.org/police_calls_for_service/pd_calls_for_service_", year, "_datasd.csv")
  read_csv(url, col_types = cols(
    incident_num = col_character(), 
    date_time = col_datetime(format = "%Y-%m-%d %H:%M:%S"), 
    day_of_week = col_integer(), 
    address_number_primary = col_integer(), 
    address_dir_primary = col_character(), 
    address_road_primary = col_character(), 
    address_sfx_primary = col_character(), 
    address_dir_intersecting = col_character(), 
    address_road_intersecting = col_character(), 
    address_sfx_intersecting = col_character(), 
    call_type = col_character(), 
    disposition = col_character(), 
    beat = col_integer(), 
    priority = col_integer()
  )) %>%
  mutate(year = year)
})

# Combine all years into a single dataframe
pd_combined <- bind_rows(pd_data_list)
```

```{r data cleaning and map}
# Combine all years into a single dataframe
pd_combined <- bind_rows(pd_data_list)

# Aggregate call counts by beat and year
crime_beat_sums <- pd_combined %>% 
  group_by(beat, year) %>% 
  summarize(call_count = n(), .groups = 'drop') %>%
  pivot_wider(names_from = year, values_from = call_count, names_prefix = "ct_")

# Load geojson data
geojson <- geojson_sf("https://seshat.datasd.org/gis_police_beats/pd_beats_datasd.geojson")

# Merge with spatial data
m <- geojson %>%
  left_join(crime_beat_sums, by = 'beat')

# Create hover text with percentage change
mytext <- apply(m, 1, function(row) {
  pct_changes <- sapply(2:length(years), function(i) {
    year_current <- as.numeric(row[paste0("ct_", years[i])])
    year_previous <- as.numeric(row[paste0("ct_", years[i-1])])
    
    if (is.na(year_previous) || year_previous == 0) {
      pct_change <- NA
    } else {
      pct_change <- (year_current - year_previous) / year_previous * 100
    }
    
    data.frame(
      year = years[i],
      calls = year_current,
      pct_change = round(pct_change, 2)
    )
  }, simplify = FALSE) %>%
    bind_rows() %>%
    arrange(desc(year))  # Arrange by year in descending order
  
  # Construct the hover text
  paste(
    "Beat: ", row["name"], "<br/>",
    "Year | Calls | % Change<br/>",
    paste(apply(pct_changes, 1, function(r) {
      paste(r["year"], r["calls"], r["pct_change"], sep = " | ")
    }), collapse = "<br/>")
  )
}) %>%
lapply(htmltools::HTML)


# Render the map with hover text
pal <- colorNumeric(palette = viridis_pal()(100), domain = m$ct_2024)

leaflet(m) %>%
  addTiles() %>%
  addPolygons(stroke = TRUE, 
              color = "black", 
              weight = 1, 
              fillOpacity = 0.5, 
              smoothFactor = 0.3, 
              label = mytext,
              highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
              fillColor = ~pal(ct_2024))

```  

```{r UI, echo=FALSE}
# Define the UI layout with the dropdown above the table
ui <- fluidPage(
  fluidRow(
    column(12,
           selectInput("selected_beat", "Select a Beat:", 
                       choices = c("ALL", unique(m$name)), selected = "ALL")
    )
  ),
  fluidRow(
    column(12,
           tableOutput("callTable")
    )
  )
)
```

```{r server, echo=FALSE}
# Define the server logic for the dropdown and table
server <- function(input, output, session) {

  # Ensure beat filtering is based on beat names in 'm' and corresponding field in 'pd_combined'
  output$callTable <- renderTable({
    if (input$selected_beat == "ALL") {
      selected_data <- pd_combined %>%
        group_by(year, call_type) %>%
        summarize(call_count = n(), .groups = 'drop') %>%
        pivot_wider(names_from = call_type, values_from = call_count) %>%
        arrange(desc(year))
    } else {
      # Filter by the selected beat name
      selected_beat_id <- m %>% filter(name == input$selected_beat) %>% pull(beat)
      
      selected_data <- pd_combined %>%
        filter(beat %in% selected_beat_id) %>%
        group_by(year, call_type) %>%
        summarize(call_count = n(), .groups = 'drop') %>%
        pivot_wider(names_from = call_type, values_from = call_count) %>%
        arrange(desc(year))
    }
    
    selected_data
  })
}

# Combine the UI and server into a Shiny app
shinyApp(ui = ui, server = server)

```


